# Regression Analysis

## SLR

::: callout-important
## Summary of Simple Linear Regression (SLR)

**1) The PRF :** $y_i =\alpha +\beta x_i +\epsilon_i$

**2) The OLS estimates of** $\alpha$ **and** $\beta$ **are:**

$$a=\bar y-b \bar x$$

$$b=\frac{n\sum xy-(\sum x)(\sum y)}{n\sum x^2 -(\sum x)^2}$$

respectively.

Hence the **estimated regression line** is:\

$$\hat y_i =a+bx$$

**3) Point prediction:** For a given value of $X$ say $x_g$ the predicted value of $Y$ is:

$$\hat y_g=a+b *(x_g)$$

**4) Goodness of fit for SLR:**

$$R^2=\frac{b^2(\sum x^2 -n \  \bar x^2)}{\sum y^2 -n \ \ \bar y^2}\ \ ; \ \ 0\le R^2\le 1$$

**5) Mean squared error (MSE): as an estimate of** $\sigma_e^2$:

$$s_e^2=\frac{\sum_{i=1}^{n} (y_i -\hat y_i)^2}{n-2}$$

**6) Hypothesis test for** $\beta$**:**

$$H_0: \beta =c \ \ vs.$$

$$H_1: \beta \ne c$$

Test statistic:

$$t_0=\frac{b-c}{s.e(b)}$$

where, $s.e(b)=\frac{s_e}{\sqrt {\sum (x-\bar x)^2}}$

Under $H_0$, $t_0$ follows *T*-*distribution with* $n-2$ df.
:::

**Problem 7.1 [@walpole2017,pp.399]:** The following is a portion of a classic data set called the “pilot plot data” in Fitting Equations to Data by Daniel and Wood, published in 1971. The response y is the acid content of material produced by titration, whereas the regressor x is the organic acid content produced by extraction and weighing.\

|       |     |     |     |     |     |     |     |     |     |     |
|:-----:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|
| **y** | 76  | 62  | 66  | 58  | 88  | 70  | 37  | 82  | 88  | 43  |
| **x** | 123 | 55  | 100 | 75  | 159 | 109 | 48  | 138 | 164 | 28  |

```{r}
# Define the function
get_slr_stats <- function(x, y) {
  # Basic counts and sums
  n <- length(x)
  sum_x <- sum(x)
  sum_y <- sum(y)
  sum_x2 <- sum(x^2)
  sum_y2 <- sum(y^2)
  sum_xy <- sum(x * y)
  
  
  x_bar <- mean(x)
  y_bar <- mean(y)
  
  slope <- (n * sum_xy - (sum_x * sum_y)) / (n * sum_x2 - (sum_x^2))
  
  intercept <- y_bar - (slope * x_bar)
  
  # Return results as a named list
  return(list(
    n = n,
    sum_x = sum_x,
    sum_y = sum_y,
    sum_x2 = sum_x2,
    sum_y2 = sum_y2,
    sum_xy = sum_xy,
    x_bar = x_bar,
    y_bar = y_bar,
    intercept = intercept,
    slope = slope
  ))
}
```

```{r}
# Creating the data frame
df <- data.frame(
  y = c(76, 62, 66, 58, 88, 70, 37, 82, 88, 43),
  x = c(123, 55, 100, 75, 159, 109, 48, 138, 164, 28)
)

# Display the data frame
#print(df)
stats <- get_slr_stats(df$x, df$y)
#print(stats)

#summary(lm(df$y~df$x))

```

## MLR
